---
title: "Insert Model vignette"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{weightedMotifAccess}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Dependencies

```{r setup}
suppressPackageStartupMessages({
    library(weightedMotifAccess)
    library(SummarizedExperiment)
    library(GenomicRanges)
    library(Rsamtools)
    library(betterChromVAR)
})
```

## Inputs

`weightedInsertions()` expects three inputs:

1. `lf`: the ATAC-seq fragments of each sample. Unlike `getWeightedCounts()`,
   which also accepts in-memory objects, `weightedInsertions()` streams over
   the genome chromosome by chromosome and therefore only accepts **paths to
   indexed files** — either indexed `.bam` files, or `bgzip`-compressed,
   tabix-indexed fragment files. The names of `lf` become the sample names.
2. `peaks`: the peak regions, as a `GRanges`. We recommend standardizing
   their width and restricting them to the standard chromosomes.
3. `motifMatches`: the motif matches, as a `GRanges` with a `motif_id`
   metadata column.

### Indexing plain fragment files

A plain, uncompressed BED/TSV file cannot be streamed. If that is what you
have, compress and index it once and pass the resulting `.gz` path:

```{r indexing, eval=FALSE}
indexed <- bgzip("sample1.fragments.bed", dest = "sample1.fragments.bed.gz")
indexTabix(indexed, format = "bed")
```

### The `shift` argument

Tn5 inserts a few bases inside each fragment end, so coordinates have to be
shifted before insertions are counted. Whether you need this depends on your
input: raw `.bam` files are unshifted and need `shift = TRUE`, whereas many
published fragment files have already been shifted by the pipeline that
produced them and need `shift = FALSE`. If you leave `shift` unspecified it
defaults to `TRUE` when all files are bams, and a message reports that it did
so. Getting this wrong displaces the footprint by a few bases, so it is worth
checking how your fragment files were generated.

## Generate dummy data

For the purpose of this example, we'll use (twice) an example bam file from the
`r BiocStyle::Biocpkg("Rsamtools")` package, and generate dummy data for the
rest:

```{r load-data}
bam <- system.file("extdata", "ex1.bam", package="Rsamtools")
bams <- c(sample1=bam, sample2=bam)
# create regions of interest
peaks <- GRanges(c("seq1","seq1","seq2"), IRanges(c(400,900,500), width=100))
motifMatches <- GRanges(c("seq1","seq1","seq2"), IRanges(c(450,950,550), width=10),
                        motif_id=c("A","A","B"))
```

On real data you would prepare the two `GRanges` roughly as below. Note in
particular that the motif matches must use the same seqlevels style as the
peaks (which in turn has to match the fragment files), and that restricting
them to the peaks up front saves a lot of work:

```{r prepare-ranges, eval=FALSE}
species <- "Homo_sapiens"
peaks <- rtracklayer::import("peaks.narrowPeak")
peaks <- keepStandardChromosomes(peaks, species = species,
                                 pruning.mode = "coarse")
peaks <- resize(peaks, width = 300, fix = "center")

motifMatches <- readRDS("motifMatches.rds")
motifMatches <- keepStandardChromosomes(motifMatches, species = species,
                                        pruning.mode = "coarse")
seqlevelsStyle(motifMatches) <- seqlevelsStyle(peaks)[1]
motifMatches <- subsetByOverlaps(motifMatches, peaks)
```

## Get the weighted insertion counts

We can then call the `weightedInsertions()` function:

```{r weightedInsertions}
data <- weightedInsertions(bams, peaks, motifMatches=motifMatches, shift=TRUE)

sapply(data, class)
```

See `?weightedInsertions` for a description of each element in the output.

## Using betterChromVAR on the weighted insertion counts

To apply `r BiocStyle::Biocpkg("betterChromVAR")` on the weighted insertion
counts, we first need to add the GC bias in the (unweighted) counts object
(i.e. `data$peakCounts`). On real data this is done with
`betterChromVAR::addGCBias()`, which needs the `BSgenome` of your organism and
the same seqlevels style as that genome:

```{r gc-real, eval=FALSE}
seqlevelsStyle(data$peakCounts) <- "UCSC"
data$peakCounts <- addGCBias(data$peakCounts,
                             genome = BSgenome.Hsapiens.UCSC.hg38)
```

Since this is dummy data on a toy genome, we'll instead set dummy values:

```{r, gc}
rowData(data$peakCounts)$bias <- rnorm(3, 0.5, 0.1)
```

We would then run a weight-friendly version of betterChromVAR's deviation
computation, filtering out any eventual peak with zero counts across samples:

```{r}
w <- which(rowSums(assay(data$peakCounts)) > 0)
dev <- betterChromVAR::computeDeviationsWeighted(
          weightedMotifCounts = data$wInsCounts,
          unweightedPeakCounts = data$peakCounts[w,],
          annotations = data$peakAnnotation[w,])
dev
```

The resulting object is a `r BiocStyle::Biocpkg("SummarizedExperiment")` object,
as produced by the other `r BiocStyle::Biocpkg("betterChromVAR")` method,
containing normalized motif deviations and z-scores based on the weighted
insertion counts.

## Putting it together

For reference, this is the whole analysis on real data:

```{r full-workflow, eval=FALSE}
library(BSgenome.Hsapiens.UCSC.hg38)
genome <- BSgenome.Hsapiens.UCSC.hg38
species <- "Homo_sapiens"

# 1. fragments: one indexed file per sample
lf <- c(sample1 = "sample1.bam", sample2 = "sample2.bam")

# 2. peaks and motif matches, on matching seqlevels
peaks <- rtracklayer::import("peaks.narrowPeak")
peaks <- keepStandardChromosomes(peaks, species = species,
                                 pruning.mode = "coarse")
peaks <- resize(peaks, width = 300, fix = "center")

motifMatches <- readRDS("motifMatches.rds")
motifMatches <- keepStandardChromosomes(motifMatches, species = species,
                                        pruning.mode = "coarse")
seqlevelsStyle(motifMatches) <- seqlevelsStyle(peaks)[1]
motifMatches <- subsetByOverlaps(motifMatches, peaks)

# 3. weighted insertion counts (shift=TRUE for unshifted bams)
res <- weightedInsertions(lf, peaks, motifMatches, shift = TRUE, ncores = 1)

# 4. GC bias on the unweighted peak counts, dropping empty peaks
pc <- res$peakCounts
seqlevelsStyle(pc) <- "UCSC"
w <- which(rowSums(assay(pc)) > 0)
pc <- addGCBias(pc[w, ], genome = genome)

# 5. motif deviations
dev <- betterChromVAR::computeDeviationsWeighted(
    res$wInsCounts, pc, annotations = res$peakAnnotation[w, ])
```

## Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
