---
title: "Insert Model vignette"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{weightedMotifAccess}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Dependencies

```{r setup}
suppressPackageStartupMessages({
    library(weightedMotifAccess)
    library(SummarizedExperiment)
    library(betterChromVAR)
})
```

## Generate dummy data

`weightedInsertions()` expects 3 inputs: 1) the filepaths to BAM or fragment 
files for each sample (`lf` argument), 2) the peak regions (as a `GRanges`, we
recommend standardizing their width and filtering for autosomal chromosomes), 
and 3) the motif matching regions.

For the purpose of this example, we'll use (twice) an example bam file from the 
`r BiocStyle::Biocpkg("Rsamtools")` package, and generate dummy data for the 
rest:

```{r load-data}
bam <- system.file("extdata", "ex1.bam", package="Rsamtools")
bams <- c(sample1=bam, sample2=bam)
# create regions of interest
peaks <- GRanges(c("seq1","seq1","seq2"), IRanges(c(400,900,500), width=100))
motifMatches <- GRanges(c("seq1","seq1","seq2"), IRanges(c(450,950,550), width=10),
                        motif_id=c("A","A","B"))
```

## Get the weighted insertion counts

We can then call the `weightedInsertions()` function:

```{r weightedInsertions}
data <- weightedInsertions(bams, peaks, motifMatches=motifMatches, shift=TRUE)

sapply(data, class)
```

See `?weightedInsertions` for a description of each element in the output.

## Using betterChromVAR on the weighted insertion counts

To apply `r BiocStyle::Biocpkg("betterChromVAR")` on the weighted insertion 
counts, we first need to add the GC bias in the (unweighted) counts object 
(i.e. `data$peakCounts`). This would normally be done with 
`betterChromVAR::addGCBias()`, but since this is dummy data, we'll set dummy 
values:

```{r, gc}
rowData(data$peakCounts)$bias <- rnorm(3, 0.5, 0.1)
```

We would then run a weight-friendly version of betterChromVAR's deviation 
computation, filtering out any eventual peak with zero counts across samples:

```{r}
w <- which(rowSums(assay(data$peakCounts)) > 0)
dev <- betterChromVAR::computeDeviationsWeighted(
          weightedMotifCounts = data$wInsCounts,
          unweightedPeakCounts = data$peakCounts[w,],
          annotations = data$peakAnnotation[w,])
dev
```

The resulting object is a `r BiocStyle::Biocpkg("SummarizedExperiment")` object,
as produced by the other `r BiocStyle::Biocpkg("betterChromVAR")` method,
containing normalized motif deviations and z-scores based on the weighted 
insertion counts.

## Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
