---
title: "Motif interaction analysis"
author: 
  - name: Pierre-Luc Germain
    affiliation:
    - ETH Zurich, Switzerland
    email: pierre-luc.germain@hest.ethz.ch
date: "`r doc_date()`"
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
package: "`r pkg_ver('betterChromVAR')`"
vignette: >
  %\VignetteIndexEntry{Motif interaction analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

## Introduction

Motif deviations, as for instance computed by the `r Biocpkg("betterChromVAR")`
package, quantify the relative accessibility of given DNA motifs across 
samples/cells. These motif scores are then use to inference the relative 
activity of the corresponding TFs, and can then be used for differential 
analysis across conditions. If you are not familiar with motif deviations, it 
is recommended that you read up on it, for instance the documentation of the 
`r Biocpkg("chromVAR")` and/or `r Biocpkg("betterChromVAR")` packages.

In this vignette, we explore the use of these methods to identify synergies 
(or antagonisms) between TFs. As a example dataset, we use a subset of an 
ATAC-seq dataset of A549 cells upon ligand-based activation of the 
glucocorticoid receptor (GCR), encoded by the NR3C1 gene.

First, we load the package and the dataset:

```{r}
devtools::load_all("../")
#data("NR3C1example.RData")
load("../data/NR3C1example.RData", verbose=TRUE)
peakSE
motifMatches
```

The `peakSE` object contains ATAC peaks counts for each 6 samples, while the 
`motifMatches` object indicates which peak matches which of a selection of 29 
motifs. For discussion about how exactly to generate those kinds of input, see
the `r Biocpkg("betterChromVAR")` package.

## Standard differential motif accessibility using betterChromVAR

First, we illustrate the effect of the treatment of GCR motif accessibility by
computing standard motif deviations, followed by differential testing
^[If you are not familiar with such differential testing, see the `r Biocpkg("limma")` documentation.]:

```{r}
library(betterChromVAR)
dev <- betterChromVAR(peakSE, motifMatches)
dev
```
```{r}
library(limma)
design <- model.matrix(~dev$group)
fit <- eBayes(lmFit(assay(dev, "z"), design))
DAA <-as.data.frame(topTable(fit))
DAA
```

The results show GCR as the most differentially-accessible motif between the 
conditions, as expected, closely followed by the mineralocorticoid receptor 
(MCR), which has a motif very similar to GCR. We could also look at the pattern
across samples for top motifs:

```{r}
library(sechm)
sechm(dev, head(row.names(DAA)), gaps_at="group")
```


## Testing for motif interactions

Having identified GCR as the main motif differentially-accessible across 
conditions, we now want to know whether these differences are particularly at 
sites that have both GCR and some other motif (or the oppsite, i.e. particularly
at sites that do not harbour the other motif). Computationally speaking, we 
want to test whether the deviation differences at intersecting sites (i.e. 
sites that harbour both motifs) are significantly larger (or smaller) than at 
GCR sites in general.

To this end, we first compute deviations at the intersection between GCR and 
each other motif:

```{r}
# get deviations for motif intersections:
idev <- getInteractionsDeviations(peakSE, motifMatches, bait="GCR", minCount = 10)
```

Next, we test for a significant interaction with respect to the bait motif:

```{r}
res <- discoverMotifInteractions(idev, "group")
head(res)
```

We could plot these results in a kind of volcano plot:

```{r}
library(ggplot2)
ggplot(res, aes(difference, -log10(adj.P.Val), size=overlap.N, label=motif)) + 
  geom_text()
```

The `difference` column indicate by how much the response in the intersection 
(e.g. in sites that have both the GCR and CTCFL motifs) differs from that in
GCR sites more generally. For example, since we established earlier that the 
treatment increases accessibility of GCR sites, and the difference in the 
interaction with CTCFL is negative, we can conclude that CTCFL appears to 
prevent GCR from having a positive impact. In contrast, sites that have the JUND
motif (in addition to GCR) appear to increase in accessibility more than those 
that do not, in line with the known interaction between GCR and JUN/FOS family 
members.

## Investigating motif interactions through distance decay patterns

Next, we will explore a bit more in depth those candidate interactions by 
investigating their dependence on the distance between the two motifs. To this
end, the `exploreMotifInteraction` function will produce deviations for the 
sites harboring only GCR, only the candidate partner, or both at given distance
intervals of each other.

```{r, fig.width=10, fig.height=4}
# we load the genome
library(BSgenome.Hsapiens.UCSC.hg38)

# we precompute the betterChromVAR background to save computing time in case
# we check many interactions:
bg <- computeBackgrounds(peakSE, getBackgroundBins(peakSE))

# We look at two potential partners of GCR; for each, we compute deviations for
# different subsets of sites, and then melt these results into a data.frame for 
# plotting
dd <- lapply(c("JUND", "CTCFL"), \(mo){
  didev <- exploreMotifInteraction(peakSE, motifs[c("GCR", mo)], bg = bg,
                                   genome=BSgenome.Hsapiens.UCSC.hg38)
  d <- meltSE(dev, row.names(didev))
  d$feature <- factor(d$feature, row.names(didev))
  d$query <- mo
  d
})
dd <- dplyr::bind_rows(dd)
dd$query <- factor(dd$query, unique(dd$query))
ggplot(dd, aes(feature, norm)) + geom_hline(yintercept=0, linetype="dashed") + 
  geom_boxplot(aes(fill=group)) + facet_wrap(~query, scale="free") + 
  theme_bw() + labs(x="Site type", y="Adjusted z-scores") + 
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
```
On the left of the first panel, we see the change in deviation upon 
ligand-based GCR activation (i.e. NR3C1 group vs CTRL) in sites harboring the 
GCR motif (independently of JUND status). Beside it, we see the same in sites 
harboring the JUND motif (independently of whether they also harbor a GCR motif)
-- and we can observe an absence of difference. Next, we see GCR sites that are
negative for JUND (which show a smaller impact that GCR sites more generally),
followed by JUND sites negative for GCR. Finally, we see sites harboring both
motifs, at difference ranges of distance from each other. The exact distances
represent quantiles (so that each distance bin is equally populated) up to a 
maximum distance specified by the user.

We see that when site have GCR and JUND motifs very close to each other, the 
difference in deviation is larger (i.e. the effect of GCR activation on 
accessibility is stronger), and that this decays with distance, as we would 
expect of a bona fide interaction. If the interaction is a physical interaction 
between the TFs, we would expect this to decay rapidly, otherwise it can decay
more slowly.

In contrast, the putative interaction (antagonism) with CTCFL shows an 
inconsistent distance pattern: CTCFL appears to suppress the effect of GCR 
activation both when close or distant to the GCR motif, but not in the middle 
bin of 220-299bp. This makes the putative interaction less likely to be a 
direct physical interaction, but does not exclude an indirect interaction.


## Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
