---
title: "Weight Model vignette"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{weightedMotifAccess}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Dependencies

```{r setup}
suppressPackageStartupMessages({
    library(weightedMotifAccess)
    library(SummarizedExperiment)
    library(GenomeInfoDb)
    library(BSgenome.Hsapiens.UCSC.hg38)
})
```

## Load data

`getWeightedCounts()` expects peak regions (as a `GRanges`) and, in the `atac`
argument, the ATAC-seq fragments of each sample. In most cases the simplest
thing to do is to point `atac` directly at your files:

```{r file-input, eval=FALSE}
se <- getWeightedCounts(
    atac = c(sample1 = "sample1.fragments.tsv.gz",
             sample2 = "sample2.fragments.tsv.gz"),
    ranges = peaks,
    genome = genome)
```

`atac` accepts, one entry per sample:

* a named character vector of **file paths** — `.bam` (paired-end; fragments
  are assembled from the read pairs and ATAC-shifted), `.bed`, `.tsv` or
  `.txt` (optionally `.gz`-compressed; the first three columns are read as
  chromosome, start and end), or `.rds`;
* a named list of **`data.table`s** (or `data.frame`s) with
  `seqnames`/`start`/`end` columns (`chr` is accepted instead of `seqnames`);
* a named list of **`GRanges`**, or a `GRangesList`.

The names become the sample names of the returned object; if you don't
provide any, they are derived from the file names. For a single sample the
object can be passed on its own, without wrapping it in a list — a lone
`data.table`, `GRanges` or file path all work.

Only the coordinate columns are used: any further columns are ignored, and
any leading `#` comment lines are skipped.

To keep this vignette self-contained and reproducible on any machine
(including Bioconductor's build servers), we don't read real files from disk:
we instead reuse the peak coordinates and sample names bundled with the
package (`NR3C1example`) and simulate fragments around them as `data.table`s.

```{r load-data}
data(NR3C1example, package = "weightedMotifAccess")
peaks <- rowRanges(peakSE)[seq_len(2000)]
genome <- BSgenome.Hsapiens.UCSC.hg38
species <- organism(genome)

makeFrags <- function(n) {
    idx <- sample(length(peaks), n, replace = TRUE)
    s <- start(peaks)[idx]
    # vary fragment widths so all default length categories (see `cuts`
    # in `?getWeightedCounts`) are represented
    widths <- rep(c(80L, 200L, 400L), length.out = n)
    data.table::data.table(
        seqnames = as.character(seqnames(peaks))[idx],
        start    = pmax(1L, s),
        end      = pmax(1L, s + widths - 1L)
    )
}
set.seed(1)
atacFrag <- setNames(
    lapply(colnames(peakSE), function(x) makeFrags(4000)),
    colnames(peakSE)
)
head(atacFrag[[1]])
```

## Filter to standard chromosomes

`getWeightedCounts()` does not filter chromosomes internally: this is left
to the caller, so you can decide exactly which chromosomes/contigs to keep
for your organism. We recommend filtering both the peaks and the fragments
to standard chromosomes before calling the function, e.g. with
`GenomeInfoDb::keepStandardChromosomes()`. (When passing file paths, only the
`peaks` half of this applies — fragments on chromosomes absent from `peaks`
are dropped internally anyway.)

```{r filter-chromosomes}
peaks <- keepStandardChromosomes(peaks, species = species,
                                  pruning.mode = "coarse")
atacFrag <- lapply(atacFrag, function(dt) {
    gr <- GenomicRanges::makeGRangesFromDataFrame(dt)
    gr <- keepStandardChromosomes(gr, species = species,
                                   pruning.mode = "coarse")
    data.table::as.data.table(gr)
})
```

## Get origin (unweighted) count matrix

```{r unweighted-counts}
se <- getWeightedCounts(atac = atacFrag,
    ranges = peaks,
    genome = genome,
    resize = TRUE)

assay(se)[1:5,1:5]
```

## Get fragWeight and peakWeight count matrix

```{r weighted-counts}
weightSE <- getWeightedCounts(atac = atacFrag,
    ranges = peaks,
    genome = genome,
    resize = TRUE,
    fragWeight = TRUE,
    peakWeight = TRUE)

assay(weightSE)[1:5,1:5]
```

## Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
